A Practical Optional Type System for Clojure

### 2012-10-25 dissertation/appendix.tex
Final commit before marking

\documentclass[12pt, a4paper]{article}

\usepackage{biblatex}
\bibliography{bibliography}

\setlength{\oddsidemargin}{0.5cm}
\setlength{\evensidemargin}{0.5cm}
\setlength{\topmargin}{-1.6cm}
\setlength{\leftmargin}{0.5cm}
\setlength{\rightmargin}{0.5cm}
\setlength{\textheight}{24.00cm} 
\setlength{\textwidth}{15.00cm}
\parindent 0pt
\parskip 5pt
\pagestyle{plain}

\title{Research Proposal}
\author{}
\date{}

\newcommand{\namelistlabel}[1]{\mbox{#1}\hfil}
\newenvironment{namelist}[1]{%1
\begin{list}{}
    {
        \let\makelabel\namelistlabel
        \settowidth{\labelwidth}{#1}
        \setlength{\leftmargin}{1.1\labelwidth}
    }
  }{%1
\end{list}}

\begin{document}
\maketitle

\begin{namelist}{xxxxxxxxxxxx}
\item[{\bf Title:}]
	An Optional Static Type System for Clojure
\item[{\bf Author:}]
	Ambrose Bonnaire-Sergeant
\item[{\bf Supervisor:}]
	Rowan Davies
\end{namelist}

@@

Tobin-Hochstadt notes that ``untyped scripts are difficult to 
maintain over the long run'' \cite{TypedScheme:2010} because types
contain valuable design information.
He developed Typed Scheme
\cite{TypedScheme:2010} 
to safely and incrementally port existing Scheme code, a dynamic language, to
Typed Scheme, a static language.

Clojure is a dynamically typed, functional language with implementations for the Java Virtual
Machine and Common Language Runtime, and compiling to Javascript. Clojure
is also a Lisp, which makes it a good candidate to test the generality of
ideas developed for Typed Scheme \cite{TypedScheme:2010}.

@@

It will be based on the lessons learnt throught the development
of Typed Scheme, and as a response to Tobin-Hochstadt's \cite{TypedScheme:2010}
suggestion to add type systems to other existing dynamic languages.

@@

Multimethods play a significant role in Clojure, and a satisfactory
type system for Clojure should understand them to some degree.
There is some experience statically typing multimethods
\cite{Millstein02modularstatically}, 
but this is a challenge and may not be addressed in the
timeframe proposed, beyond initial consideration of how it would
fit with the rest of the system.

Clojure's core library includes variable-arity functions which
are not easily typed with current static type systems. 
Providing satisfactory types for functions 
like Clojure's 'map', 'filter', and 'reduce' require
support for non-uniform variable arity polymorphism. 
Strickland, Tobin-Hochstadt and Felleisen \cite{Strickland:2009:PVP:1532974.1532978}.
describe their approach for non-uniform variable arity polymorphism, as used
in Typed Racket. 
At a minimum, the proposed project will include broadly identifiying 
what new issues arise when adapting this approach to Clojure, 
with a prototype design if these issues are not too major.

Occurence typing is type checking technique developed for Typed Scheme 
\cite{Tobin-Hochstadt:2008:DIT:1328897.1328486}
and improved for Typed Racket
\cite{Tobin-Hochstadt:2010:LTU:1932681.1863561}.
It helps the type checker understand common programming idioms 
with minimal type annotations.
The proposed project will include a comparison of these approaches in
the context of Clojure, and a prototype design if a satisfactory approach 
is designed and no major issues are identifed.

Ensuring vigorous type safety is an important aspect of Typed Scheme 
\cite{TypedScheme:2010} ,
especially when interacting between untyped and typed modules.
I do not expect to concentrate on every combination of cross-module
interaction. In particular, safely using typed code from untyped code
will not be designed. Safe interaction between typed modules
however is needed for a practical type checker. I
will attempt to design a viable strategy or identify issues that
require further consideration.

>>>>

\appendix
\chapter{Heterogeous map type theory prototype}
\label{sec:hmaptheory}

\section{Operational Semantics}

%$$
%\begin{tdisplay}{Test1}
%\begin{altgrammar}
%  \s{},\t{} &::= & \int \alt \bool \alt {\proctype {\s{}} {\t{}}} 
%  &\mbox{Types}\\
%  \M{}, \N{}, \P{}, \dots &::=& \x{} \alt \abs{\x{}}{\M{}} \alt
%  \comb{\M1}{\M2}  &\mbox{Raw Terms} \\
%\end{altgrammar}
%\end{tdisplay} 
%$$

$$
\begin{tdisplay}{Syntax of Terms}
\begin{altgrammar}
  \exp{} &::=& (\c{} \overrightarrow{\exp{}})         % function application, multiple arguments
             \alt \{ \overrightarrow{\exp{}\ \exp{}} \} % map literal
             &\mbox{Expressions} \\ 
  \v{} &::=& \k{} \alt \nil{}
              \alt \{ \overrightarrow{\v{}\ \v{}} \}   % map literal
              &\mbox{Values} \\
  \c{} &::=& \assoc{} \alt \dissoc{} \alt \get{}
              &\mbox{Constants}
\end{altgrammar}
\end{tdisplay}
$$

$$
\begin{tdisplay}{Evaluation Contexts}
  \begin{altgrammar}
    \E{} &::=& [ ] % application rules
              \alt (\c{}\ \overrightarrow{\v{}}\ \E{}\ \overrightarrow{\exp{}}) % eval arguments left-to-right
              % map rules
              \alt \{\overrightarrow{\v{}\ \v{}}\ \E{}\ \exp{}\ \overrightarrow{\exp{}\ \exp{}} \} % key first
              \alt \{\overrightarrow{\v{}\ \v{}}\ \v{}\ \E{}\ \overrightarrow{\exp{}\ \exp{}} \}   % value next
              &\mbox{Evaluation Contexts}
  \end{altgrammar}
\end{tdisplay}
$$

% \mapsto for updating maps
$$
\begin{tdisplay}{Operational Semantics}
\begin{array}{c}
{\inferrule[E-Assoc]
  { {v_1 = \{ \overrightarrow{\v{}\ \v{}} \} } \\
    {v_4 = v_1\ with\ entry\ v_2\ to\ v_3} }
    {{(\assoc{}\ v_1\ v_2\ v_3) \hookrightarrow v_4} }} \\
\\
{\inferrule[E-Dissoc]
  { {v_1 = \{ \overrightarrow{\v{}\ \v{}} \} } \\
    {v_3 = v_1\ without\ entry\ indexed\ by\ v_2} }
  {(dissoc\ v_1\ v_2) \hookrightarrow v_3} } \\
\\
{\inferrule[E-GetMapExist]
  { {v_1 = \{ \overrightarrow{\v{}\ \v{}} \} } \\
    {v_2\ v_3\ in\ v_1} }
%    {v_1\ has\ entry\ with\ key\ v_2} \\
%    {v_3 = corresponding\ value\ in\ entry\ with\ key\ v_2\ in\ v_1}
  { {(get\ v_1\ v_2) \hookrightarrow v_3} }} \\
\\
{\inferrule[E-GetMapNotExist]
  { {v_1 = \{ \overrightarrow{\v{}\ \v{}} \} } \\
    {v_1\ has\ no\ entry\ with\ key\ v_2} \\
    {v_3 = \nil{}} }
  { {(get\ v_1\ v_2) \hookrightarrow v_3} }}
\end{array} 
\end{tdisplay} 
$$

$$
\begin{tdisplay}{Type Syntax}
\begin{altgrammar}
  \T{} &::=& \quad \Top \alt \Bottom \alt \Nil \alt \k{} \alt \{ \overrightarrow{\k{}\ \T{}} \} 
             \alt {\IPersistentMap {\T{}} {\T{}}} \\
             && \alt ( \vee\ \overrightarrow{\T{}} )
                \alt ( \wedge\ \overrightarrow{\T{}} )
\end{altgrammar}
\end{tdisplay}
$$

% what about get? Core Type Rules

$$
\begin{tdisplay}{Core Type Rules}
\begin{array}{c}
{\inferrule[T-AssocHMap]
  { {\Gamma \vdash {\hastype {{\exp{}}_{m}} {{\T{}}_{m}}} } \\
    {\Gamma \vdash {\hastype {{\exp{}}_{k}} {{\k{}}}} } \\
    {\Gamma \vdash {\hastype {{\exp{}}_{v}} {{\T{}}_{v}}} } \\
    {\vdash {\issubtype {{\T{}}_{m}} {\TopHMap} }} }
  { {\Gamma \vdash {\hastype {( \assoc{}\ {\exp{}}_{m}\ {\exp{}}_{k}\ {\exp{}}_{v})} 
        {(update\_hmap\ {{\T{}}_{m}}\ {{\k{}}}\ {{\T{}}_{v}})}}}}} \\
\\
{\inferrule[T-AssocPromote]
  { {\Gamma \vdash {\hastype {{\exp{}}_{m}} {{\T{}}_{m}}} } \\
    {\Gamma \vdash {\hastype {{\exp{}}_{k}} {{\T{}}_{k}}} } \\
    {\Gamma \vdash {\hastype {{\exp{}}_{v}} {{\T{}}_{v}}} } \\
    {\vdash {\issubtype {{\T{}}_{m}} {\TopIPersistentMap}}} \\
    {\T{} = (promote\_hmap\ {\T{}}_{m}\ {\T{}}_{k}\ {\T{}}_{v})} }
  { {\Gamma \vdash {\hastype {( \assoc{}\ {\exp{}}_{m}\ {\exp{}}_{k}\ {\exp{}}_{v})} {\T{}}}}}} \\
\end{array}
\end{tdisplay}
$$

$$
\begin{tdisplay}{Type Metafunctions}
\begingroup
\fontsize{10pt}{12pt}
\begin{altgrammar}
  (update\_hmap\ \{ \overrightarrow{{\T{}}_{k}\ {\T{}}_{v}} \}\ {\k{}}_1\ {\T{}}_1) = 
                 \{ \overrightarrow{{\T{}}_{k}\ {\T{}}_{v}}\ {\k{}}_1\ {\T{}}_1 \}\ \\
  (update\_hmap\ (\wedge\ \overrightarrow{{\T{}}})\ {\T{}}_{k}\ {\T{}}_{v}) = 
                               (\wedge\ \overrightarrow{(update\_hmap\ {\T{}}\ {\T{}}_{k}\ {\T{}}_{v})}) \\
  (update\_hmap\ (\vee\ \overrightarrow{{\T{}}})\ {\T{}}_{k}\ {\T{}}_{v}) = 
                               (\vee\ \overrightarrow{(update\_hmap\ {\T{}}\ {\T{}}_{k}\ {\T{}}_{v})})
  \\ \\
  (promote\_hmap\ \{ \overrightarrow{{\T{}}_{k}\ {\T{}}_{v}} \}\ {\T{}}_{kn}\ {\T{}}_{vn})
                                    = {\IPersistentMap {(\vee\ \overrightarrow{{\T{}}_{k}}\ {\T{}}_{kn})}
                                                       {(\vee\ \overrightarrow{{\T{}}_{v}}\ {\T{}}_{vn})}} \\
  (promote\_hmap\ {\IPersistentMap {{\T{}}_{k}} {{\T{}}_{v}}} \ {\T{}}_{kn}\ {\T{}}_{vn})
                                    = {\IPersistentMap {(\vee\ {\T{}}_{k}\ {\T{}}_{kn})}
                                                       {(\vee\ {\T{}}_{v}\ {\T{}}_{vn})}}
\end{altgrammar}
\endgroup
\end{tdisplay}
$$

%$$
%\begin{tdisplay}{Subtyping}
%\begin{array}{c}
%\inferrule[S-Refl] { } {\vdash {\issubtype {\T{}} {\T{}}}} \\
%\inferrule[S-Any] { } {\vdash {\issubtype {\T{}} {\Top}}}
%\inferrule[S-UnionSuper] 
%  {\ } 
%  {\vdash {\issubtype {\T{}} {\Top}}}
%\end{array}
%\end{tdisplay}
%$$

\chapter{Dissertation Proposal}

@@

Tobin-Hochstadt notes that ``untyped scripts are difficult to 
maintain over the long run''~\cite{Tob10} because types
contain valuable design information.
He developed Typed Scheme~\cite{Tob10} 
to safely and incrementally port existing Scheme code, a dynamic language, to
Typed Scheme, a static language.

Clojure is a dynamically typed, functional language with implementations for the Java Virtual
Machine and Common Language Runtime, and compiling to Javascript. Clojure
is also a Lisp, which makes it a good candidate to test the generality of
ideas developed for Typed Scheme~\cite{Tob10}.

@@

It will be based on the lessons learnt throught the development
of Typed Scheme, and as a response to Tobin-Hochstadt's~\cite{Tob10}
suggestion to add type systems to other existing dynamic languages.

@@

Multimethods play a significant role in Clojure, and a satisfactory
type system for Clojure should understand them to some degree.
There is some experience statically typing multimethods~\cite{MS02}, 
but this is a challenge and may not be addressed in the
timeframe proposed, beyond initial consideration of how it would
fit with the rest of the system.

Clojure's core library includes variable-arity functions which
are not easily typed with current static type systems. 
Providing satisfactory types for functions 
like Clojure's 'map', 'filter', and 'reduce' require
support for non-uniform variable arity polymorphism. 
Strickland, Tobin-Hochstadt and Felleisen~\cite{STF09}.
describe their approach for non-uniform variable arity polymorphism, as used
in Typed Racket. 
At a minimum, the proposed project will include broadly identifiying 
what new issues arise when adapting this approach to Clojure, 
with a prototype design if these issues are not too major.

Occurence typing is type checking technique developed for Typed Scheme~\cite{TF08}
and improved for Typed Racket~\cite{TF10}.
It helps the type checker understand common programming idioms 
with minimal type annotations.
The proposed project will include a comparison of these approaches in
the context of Clojure, and a prototype design if a satisfactory approach 
is designed and no major issues are identifed.

Ensuring vigorous type safety is an important aspect of Typed Scheme~\cite{Tob10},
especially when interacting between untyped and typed modules.
I do not expect to concentrate on every combination of cross-module
interaction. In particular, safely using typed code from untyped code
will not be designed. Safe interaction between typed modules
however is needed for a practical type checker. I
will attempt to design a viable strategy or identify issues that
require further consideration.

### 2012-10-25 dissertation/lit-review.tex
Add proposal, acknowledgements
Final commit before marking

Typed Clojure adds static type checking to an existing untyped language.
Coming from the other direction, Abadi, Cardelli, Pierce and R\'{e}my~\cite{ACPR94}
extend a static language to be more dynamic.
This work centres around adding a type \emph{Dynamic}, which represents
a dynamic type in a statically typed language.

@@

Strickland et al.\ invented a type system supporting variable-arity polymorphism~\cite{STF09}
a version of which is included in the current implementation of Typed Racket.
Their main innovation centres around \emph{dotted type variables}, which represent a heterogeneous sequence
of types. Dotted type variables allow \emph{non-uniform} variable-arity function types,
which are used to check definitions and usages of functions with non-trivial rest parameters

>>>>

Typed Clojure adds static type checking to an existing untyped language.
Coming from the other direction, Abadi, Cardelli, Pierce and R\'{e}my~\cite{ACPR95}
extend a static language to be more dynamic.
This work centres around adding a type \emph{Dynamic}, which represents
a dynamic type in a statically typed language.

@@

Strickland et al.\ invented a type system supporting variable-arity polymorphism ~\cite{STF09}
a version of which is included in the current implementation of Typed Racket.
Their main innovation centres around \emph{dotted type variables}, which represent a heterogeneous sequence
of types. Dotted type variables allow \emph{non-uniform} variable-arity function types,
which are used to check definitions and usages of functions with non-trivial rest parameters

### 2012-10-25 dissertation/paper.tex
Final commit before marking

I'd also like to thank: Sam Tobin-Hochstadt for his consistent assistance and patience;
David Nolen for mentoring this project as part of Google Summer of Code 2012;
and Chas Emerick, without whom I would have never considered attending Clojure Conj 2011
(and this work probably would not have happened).
Also, Daniel Spiewak, Dan Friedman and William Byrd, Jim Duey, Baishampayan Ghose, and
Steve Strickland, thank you for your help and inspiration.
Thank you Rich Hickey for creating Clojure, it is a thrill to build tools that make
this great language even better.

>>>>

I'd also like to thank: Sam Tobin-Hochstadt for his consistent assistance and patience;
David Nolen for mentoring this project as part of Google Summer of Code 2012;
and Chas Emerick, without whom I would have never considered attending Clojure Conj 2011
(and this work probably would not have happened).
Also, Daniel Spiewak, Dan Friedman and William Byrd, Jim Duey, Baishampayan Ghose, Sam Aaron, and
Steve Strickland, thank you for your help and inspiration.
Thank you Rich Hickey for creating Clojure, it is a thrill to build tools that make
this great language even better.
